Recognizing emotions with technology and starting discussions

### 2018-01-21 paper.tex
Add Material section
The material section explains the technical details of the prototype

\section{Methods}
To be added

@@

\section{Acknowledgment}
This research was supported by the Amsterdam University of Applied Sciences, the Waag Society research
institute, and Research Group Crossmedia. We thank Bernadette Schrandt for all the assistance
with the research, including guidance and feedback. We thank Lodewijk Loos and Douwe-Sjoerd Boschman 
from Waag Society for the assistance during the creation of the prototype. We thank Wouter Meys for setting
up the minor and assisting with the research. Lastly, we would like to thank all the participants who helped
us test our final prototype.

>>>>

\section{Material}
Before creating the prototype, we were free in choosing \emph{how} we were going to implement the emotion
recognition. However, there were a couple restrictions and requirements. First, a very limited time span
to create the prototype in. The prototype had to be up and running in roughly 8 weeks so that we could
use it to perform our research. Secondly, preferable everything about the prototype had to be open source, as
requested by the Waag Society. Because of prior desk-research, combined with the above restrictions, we
chose to use facial expression recognition using a camera as the basis for our prototype. This decision was
further solidified upon the discovery of several articles regarding open source facial expression software
using Python, OpenCV, Dlib, and Facial Landmarks \cite{gent2016landmarks}.

Aside from the facial expression recognition part of the project, the prototype had to be able to visualize the
emotions to allow participants to reflect on their emotions and spark a discussion. The goal of the
visualization is to show the user of the prototype their emotions over a certain time span, and allow the user
to navigate through this time span. One of our clients, the Waag Society, had already done work regarding
this subject, including the classification of emotions \cite{plutchik1980general} which we will use, 
and a possible model for the visualization. We have used both the classification, as well as the model
as a basis for our prototype. The emotions we have used in our prototype are as follows: \emph{sadness, 
anger, contempt, disgust, fear, happiness, neutral, and surprise}.

In order to support our user research, the prototype has to go through a number of steps. Below, each of these
steps is explained in more detail:
\begin{enumerate}
    \item{Show a video to the research subject, and start filming at the same time. Store the video for analysis.}
    \item{Once the video is done, analyze the video and store the results of the analysis.}
    \item{Visualize the generated model on-screen when the user is ready.}
\end{enumerate}

The first step, recording and storing the video, is done with OpenCV\footnote{\url{https://opencv.org}} 
to allow video capture from the webcam. The prototype records at 10 frames per second (FPS).
Next, the video captures are stored locally on disk.
Displaying the video that the user of the prototype watches while he is being recorded, is considered part of
the visualization, and will be discussed later in this chapter.

The analysis and prediction of emotion from video frames is largely the same as described in the series of
articles written by Paul van Gent \cite{gent2016landmarks}. We use a trained, linear SVM machine learning model
to predict the emotions on someone's face, based on 68 specific facial landmarks. The linear SVM machine learning
model was created with scikit-learn\footnote{\url{https://scikit-learn.org}}, an open source machine learning library
for Python. In order to train the SVM model, we needed a suitable dataset of images, labeled with the
accompanying emotions. For this, we used the Cohn-Kanade and Extended Cohn-Kanade 
\cite{kanade2000comprehensive, lucey2010extended} data sets.
Each image in the dataset is preprocessed before being used to train the SVM model. This includes detecting
a face on each image, cropping and resizing the face to a common size, removing all color, and using OpenCV's
CLAHE algorithm. Once the preprocessing and sorting of the images is done, we use Dlib's
\footnote{\url{https://dlib.net}} shape predictor to fetch the 68 specific facial landmarks. These landmarks,
together with the label describing the emotion on the face, are then used to train the SVM model. The total
dataset used to train the model contains 834 images. Using a larger dataset will likely result in a more accurate,
stable, result.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.45\textwidth, scale=1]{landmarks.png}
    \caption{68 facial landmarks}
\end{figure}

The recorded video from our prototype is analyzed frame by frame. Each frame goes through the same preprocessing
steps as the dataset used to train the SVM model. Once preprocessed, the 68 landmarks from each frame are then
sent through the SVM model. Since this is a probabilistic model, the percentage chance for each of the 8 emotions
is returned as a result. We found that these results could fluctuate fairly rapidly in a short number of frames.
In order to get a more stable result, we've used the average of 5 each emotion over a span of 5 frames (half a
second). This means that our final prototype predicts the emotions for every half a second of recorded video.
Once the predictions are made for the entire video, the results are stored in an Sqlite database to be used
later for the visualization. This also allows us to store the results for analysis later in our research.

The visualization itself is a web application. A Flask\footnote{\url{http://flask.pocoo.org}} webserver 
is used to allow API access to the rest of the prototype. HTML, CSS and Javascript are then used to 
create the user interface for the prototype, including the visualization of the model.
Initially, we wanted to use Qt\footnote{\url{https://qt.io}}, a C++ UI library, but due to a lack of time and
expertise on our side, we decided to opt for something more familiar, which in this case resulted in a web
application. The user interface consists of four pages. One to display the video (and start the recording),
one to manage the sessions with our research participants, one to create a new session, and lastly, one to
show the visualization. The visualization page uses doughnut charts for each recorded participants. The charts
are generated using Chart.js\footnote{\url{http://chartjs.org}}, a Javascript library for creating a multitude
of charts. The predicted emotions are displayed half a second at a time, with a time line slider to allow the
participant to scroll through the predictions throughout the video.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.45\textwidth, scale=1]{visualization.png}
    \caption{Visualized model as generated by the prototype}
\end{figure}

\section{Method}
To be added

@@

\section{Acknowledgment}
This research was supported by the Amsterdam University of Applied Sciences, the Waag Society research
institute, and Research Group Crossmedia. We thank Bernadette Schrandt for all the assistance
with the research, including guidance and feedback. We thank Lodewijk Loos and Douwe-Sjoerd Boschman 
from Waag Society for the assistance during the creation of the prototype. We thank Wouter Meys for setting
up the minor and advice with the research. Lastly, we would like to thank all the participants who helped
us test our final prototype.
